Quantifying and controlling biases in dark matter halo concentration estimates

### 2016-08-25 paper/ms.tex
updated caption

\begin{figure*}
\begin{center}
  \includegraphics[width=0.49\textwidth]{avg_foff_bolshoi.pdf}
  \includegraphics[width=0.48\textwidth]{sigma_foff_bolshoi.pdf}
\end{center}
\vspace{-0.5cm}
\caption{\emph{Left panel}. Average value of the fractional offset ($f_{\rm
    off}$ (Eq. \ref{eq:f_off}) between the concentration in a halo at 
  high resolution and its bootstrapped versions. 
  Thick (thin) lines correspond to massive halos in the Bolshoi (Via
  Lactea) simulation.
  The density method noticeably overestimates the concentration up to
  a factor of $20\%$, 
  while the new method only underestimates the concentrations by less
  than $3\%$.
  \emph{Right panel.}
  1$\sigma$ uncertainties on the offset and the concentration
  estimates from the integrated mass method.  
  Lines show the width between the 14 and 86 percentiles
  of the $f_{\rm off}$ distribution at fixed particle number. 
  The lines include the results for the three methods: maximum
  velocity, density and integrated mass.
  Grey symbols correspond to the $1\sigma$ uncertainty on each of
  the halos in the Bolshoi Simulation as a result of the MCMC
  sampling. 
  To allow a fair comparison with $f_{\rm off}$, this uncertainty has
  been normalized by the preferred concentration value for each halo. 
  \label{fig:downsampling}}
\end{figure*}

@@

Using bootstrapping data (\S \ref{sec:bootstrapping}) we find that at
a given normalized radius, $x$, the logarithm of the normalized
integrated mass, $m$, approximately follows a gaussian distribution
with variance

@@

However, tests on the bootstrapping data show that using
$\sigma_i^2=\sigma_x^2(x_i)$, instead of the full inverse covariance
matrix, grossly overestimates $\chi^2(c,x_i)$,
providing small uncertainties around the best concentration value. 
To avoid the expensive computation and inversion of a full covariance
matrix we use the bootsrapping data to calibrate an effective
$\sigma_i^2\approx \sigma_{\rm eff}^2(x_i)$.

@@

Runtime is roughly proportional to $N$. Using a single 2.3Ghz CPU core with two
walkers over 500 steps takes $\sim 0.5$ miliseconds per halo per
particle in the halo, i.e. a halo with $N=2\times 10^{3}$ can be fit
in one second.

@@

The circles in the same Figure show the $1\sigma$ uncertainty on all
the relaxed halos in the Bolshoi simulation sample using the MCMC
results.  
To allow for a fair comparison with the bootstrapping results, this
uncertainty is normalized to the concentration value.
The uncerainty from the bootstrapping experiment provides an upper
bound uncertainty on the concentration estimate for individual halos.

>>>>

\begin{figure*}
\begin{center}
  \includegraphics[width=0.49\textwidth]{avg_foff_bolshoi.pdf}
  \includegraphics[width=0.48\textwidth]{sigma_foff_bolshoi.pdf}
\end{center}
\vspace{-0.5cm}
\caption{\emph{Left panel}. Bias estimated via bootstrapping on the
  concentration as a function of particle number.
  Thick (thin) lines correspond to massive halos in the Bolshoi (Via
  Lactea) simulation.
  The density method noticeably overestimates the concentration up to
  a factor of $20\%$, 
  while the new method only underestimates the concentrations by less
  than $3\%$.
  \emph{Right panel.}
  1$\sigma$ uncertainties on the bootrsapped halos (lines) and the MCMC
  uncertainties on the concentration estimates for each halo using the
  integrated mass method (circles).   
  Lines show the width between the 14 and 86 percentiles
  of the $f_{\rm off}$ distribution at fixed particle number. 
  The lines include the results for the three methods using Bolshoi data.
  To allow a fair comparison agains $f_{\rm off}$, the MCMC uncertainty has
  been normalized by the preferred concentration value for each halo. 
  \label{fig:downsampling}}
\end{figure*}

@@

Using bootstrapping data (\S \ref{sec:bootstrapping}) we find that at
a given normalized radius, $x$, the logarithm of the normalized
integrated mass, $m$, approximately follows a Gaussian distribution
with variance

@@

However, tests on the bootstrapping data show that using
$\sigma_i^2=\sigma_x^2(x_i)$, instead of the full inverse covariance
matrix, grossly overestimates $\chi^2(c,x_i)$,
providing small uncertainties around the best concentration value. 
To avoid the expensive computation and inversion of a full covariance
matrix we use the bootstrapping data to calibrate an effective
$\sigma_i^2\approx \sigma_{\rm eff}^2(x_i)$.

@@

Run-time is roughly proportional to $N$. Using a single 2.3Ghz CPU core with two
walkers over 500 steps takes $\sim 0.5$ milliseconds per halo per
particle in the halo, i.e. a halo with $N=2\times 10^{3}$ can be fit
in one second.

@@

The circles in the same Figure show the $1\sigma$ uncertainty on all
the relaxed halos in the Bolshoi simulation sample using the MCMC
results.  
To allow for a fair comparison with the bootstrapping results, this
uncertainty is normalized to the concentration value.
The uncertainty from the bootstrapping experiment provides an upper
bound uncertainty on the concentration estimate for individual halos.
